---
title: "Differential treatment response of predicted endotypes of MEDIVIR"
date: "`r Sys.Date()`"
toc-title: "Contents"
author: "MTH (mth@nordicbio.com)"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: false
    theme: united
---

<style>
body {
text-align: justify}
</style>

```{r setup, include = FALSE, fig.align = "center"}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(stringr)
library(glmnet)
library(caret)
library(emmeans)
library(forestplot)
options(dplyr.summarise.inform = FALSE)
"%ni%" = Negate("%in%")
```



# Introduction

Assess differential treatment response in predicted endotypes in MEDIVIR using the trained model from APPROACH.

```{r}
# Fetch predicted endotype labels
wkdir = "J:/16_GITHUB/MTH/IM_MTH_MEDIVIR/endotype_prediction/"
filename = "predicted.endotypes.baseline.tsv"
predicted.endotypes = read.table(file = paste0(wkdir, filename),
                                 header = TRUE)

# Get clinical information
wkdir = "R:/STATS/02 WP/WP001_RHEUMA/ST016_Medivir/data_gdpr/DATA_COLLECTED_2024/"
filename = "medivir_clinical_30OCT2024.tsv"
data.clinical = read.table(file = paste0(wkdir, filename),
                  header = TRUE, 
                  sep = "\t") %>%
  rename(SUBJ_ID = SUBJECT) %>%
  mutate(VISIT = paste0("VISIT", VISIT)) %>%
  pivot_longer(cols = contains(c("WOMAC", "NRS")),
               names_to = "parameter") %>%
  pivot_wider(names_from = c("VISIT", "parameter"),
              values_from = value)

# Get sCXT-I and uCTX-II measurements (secondary outcome measures from study)
wkdir = "R:/STATS/02 WP/WP001_RHEUMA/ST016_Medivir/data_gdpr/DATA_COLLECTED_2024/"
filename = "medivir_biomarkers_RF_imputed_30OCT2024.tsv"
data.biomarkers = read.table(file = paste0(wkdir, filename),
                             header = TRUE) %>%
  rename(SUBJ_ID = SUBJECT) %>%
  select(SUBJ_ID, contains("CTX")) %>%
  # Back-transform measurements
  mutate(across(-SUBJ_ID, ~exp(.) - 1))

data = predicted.endotypes %>% 
  left_join(data.clinical, by = "SUBJ_ID") %>%
  left_join(data.biomarkers, by = "SUBJ_ID") %>%
  # Calculate changes
  mutate(WOMACPAIN_CH = VISIT8_WOMAC_PAIN - VISIT2_WOMAC_PAIN,
         WOMACFUNC_CH = VISIT8_WOMAC_FUNC - VISIT2_WOMAC_FUNC,
         WOMACSTIFF_CH = VISIT8_WOMAC_STIFF - VISIT2_WOMAC_STIFF,
         NRSKNEEPAIN_CH = VISIT8_NRS_KNEE_PAIN - VISIT2_NRS_KNEE_PAIN,
         # Change biomarkers to percent change from baseline
         SCTX_CH = 100*(VISIT8_sCTX_I - VISIT2_sCTX_I)/VISIT2_sCTX_I,
         UCTXII_CH = 100*(VISIT8_uCTX_II_CREA - VISIT2_uCTX_II_CREA)/VISIT2_uCTX_II_CREA,
         # Add numerical differences in biomarkers as well
         SCTX_CH_NUM = VISIT8_sCTX_I - VISIT2_sCTX_I,
         UCTXII_CH_NUM = VISIT8_uCTX_II_CREA - VISIT2_uCTX_II_CREA) %>%
  mutate_at(vars(predicted_endotype, Sex), factor)

data$TREATMENT_SHORT = factor(data$TREATMENT_SHORT, levels = c("PLACEBO", "MIV-711"))
data$predicted_endotype = relevel(data$predicted_endotype, ref = "structural_damage")
```



# Treatment response - ANCOVA
## NRS Knee pain

Primary study outcome was change from visit 2 to 8 in NRS average target knee pain score.

```{r}
lm.fit = lm(NRSKNEEPAIN_CH ~ TREATMENT_SHORT + Age + Sex + BMI + TREATMENT_SHORT*predicted_endotype + VISIT2_NRS_KNEE_PAIN, data = data)
em = emmeans(lm.fit, ~TREATMENT_SHORT|predicted_endotype)
pairs(em, rev = TRUE) %>% rbind(adjust = "holm")
```

Get confidence intervals.

```{r}
res = summary(pairs(em, rev = TRUE) %>% rbind(adjust = "holm"))

# Structural damage
str.estimate.1 = res$estimate[1]
se = res$SE[1]
df = res$df[1]
str.lower.1 = str.estimate.1 - qt(p = 0.975, df = df)*se
str.upper.1 = str.estimate.1 + qt(p = 0.975, df = df)*se

# Inflammatory
infl.estimate.1 = res$estimate[2]
se = res$SE[2]
df = res$df[2]
infl.lower.1 = infl.estimate.1 - qt(p = 0.975, df = df)*se
infl.upper.1 = infl.estimate.1 + qt(p = 0.975, df = df)*se

# Low turnover
low.estimate.1 = res$estimate[3]
se = res$SE[3]
df = res$df[3]
low.lower.1 = low.estimate.1 - qt(p = 0.975, df = df)*se
low.upper.1 = low.estimate.1 + qt(p = 0.975, df = df)*se
```


Overall test if there is difference in effect (remove interaction term in simpler model).

```{r}
lm.fit2 = lm(NRSKNEEPAIN_CH ~ TREATMENT_SHORT + Age + Sex + BMI + VISIT2_NRS_KNEE_PAIN, data = data)
anova(lm.fit2, lm.fit)
```

Overall effect.

```{r}
em = emmeans(lm.fit2, ~TREATMENT_SHORT)
pairs(em, rev = TRUE) %>% rbind(adjust = "holm")
```

```{r}
res = summary(pairs(em, rev = TRUE) %>% rbind(adjust = "holm"))

# Overall effect
overall.estimate.1 = res$estimate
se = res$SE
df = res$df
overall.lower.1 = overall.estimate.1 - qt(p = 0.975, df = df)*se
overall.upper.1 = overall.estimate.1 + qt(p = 0.975, df = df)*se
```

```{r, fig.width = 5}
dat.forest = tibble(mean = c(str.estimate.1, infl.estimate.1, low.estimate.1, overall.estimate.1),
       lower = c(str.lower.1, infl.lower.1, low.lower.1, overall.lower.1),
       upper = c(str.upper.1, infl.upper.1, low.upper.1, overall.upper.1),
       endotype = c("Structural damage", "Inflammatory", "Low tissue turnover", "Overall"))

dat.forest %>%
  forestplot(labeltext = endotype,
             xlab = "Delta NRS Knee pain (%)",
             ci.vertices = TRUE,
             lineheight =  unit(8.6,"mm"),
             boxsize = 0.25,
             col = fpColors(zero = "black")) |>
  fp_add_header(endotype = "Endotype") |>
  fp_set_style(box = "darkblue",
               line = "royalblue",
               txt_gp = fpTxtGp(xlab = gpar(cex = 1.5),
                                ticks = gpar(cex = 1.5),
                                cex = 1.5))
```



## WOMAC Pain

```{r}
lm.fit = lm(WOMACPAIN_CH ~ TREATMENT_SHORT + Age + Sex + BMI + TREATMENT_SHORT*predicted_endotype + VISIT2_WOMAC_PAIN, data = data)
em = emmeans(lm.fit, ~TREATMENT_SHORT|predicted_endotype)
pairs(em, rev = TRUE) %>% rbind(adjust = "holm")
```

Get confidence intervals.

```{r}
res = summary(pairs(em, rev = TRUE) %>% rbind(adjust = "holm"))

# Structural damage
str.estimate.2 = res$estimate[1]
se = res$SE[1]
df = res$df[1]
str.lower.2 = str.estimate.2 - qt(p = 0.975, df = df)*se
str.upper.2 = str.estimate.2 + qt(p = 0.975, df = df)*se

# Inflammatory
infl.estimate.2 = res$estimate[2]
se = res$SE[2]
df = res$df[2]
infl.lower.2 = infl.estimate.2 - qt(p = 0.975, df = df)*se
infl.upper.2 = infl.estimate.2 + qt(p = 0.975, df = df)*se

# Low turnover
low.estimate.2 = res$estimate[3]
se = res$SE[3]
df = res$df[3]
low.lower.2 = low.estimate.2 - qt(p = 0.975, df = df)*se
low.upper.2 = low.estimate.2 + qt(p = 0.975, df = df)*se
```

Overall test if there is difference in effect (remove interaction term in simpler model).

```{r}
lm.fit2 = lm(WOMACPAIN_CH ~ TREATMENT_SHORT + Age + Sex + BMI + VISIT2_WOMAC_PAIN, data = data)
anova(lm.fit2, lm.fit)
```

Overall effect.

```{r}
em = emmeans(lm.fit2, ~TREATMENT_SHORT)
pairs(em, rev = TRUE) %>% rbind(adjust = "holm")
```

```{r}
res = summary(pairs(em, rev = TRUE) %>% rbind(adjust = "holm"))

# Overall effect
overall.estimate.2 = res$estimate
se = res$SE
df = res$df
overall.lower.2 = overall.estimate.2 - qt(p = 0.975, df = df)*se
overall.upper.2 = overall.estimate.2 + qt(p = 0.975, df = df)*se
```

```{r, fig.width = 5}
dat.forest = tibble(mean = c(str.estimate.2, infl.estimate.2, low.estimate.2, overall.estimate.2),
       lower = c(str.lower.2, infl.lower.2, low.lower.2, overall.lower.2),
       upper = c(str.upper.2, infl.upper.2, low.upper.2, overall.upper.2),
       endotype = c("Structural damage", "Inflammatory", "Low tissue turnover", "Overall"))

dat.forest %>%
  forestplot(labeltext = endotype,
             xlab = "Delta WOMAC Pain (%)",
             ci.vertices = TRUE,
             boxsize = 0.25,
             lineheight =  unit(8.6,"mm"),
             col = fpColors(zero = "black")) |>
  fp_add_header(endotype = "Endotype") |>
  fp_set_style(box = "darkblue",
               line = "royalblue",
               txt_gp = fpTxtGp(xlab = gpar(cex = 1.5),
                                ticks = gpar(cex = 1.5),
                                cex = 1.5))
```



## WOMAC Function

```{r}
lm.fit = lm(WOMACFUNC_CH ~ TREATMENT_SHORT + Age + Sex + BMI + TREATMENT_SHORT*predicted_endotype + VISIT2_WOMAC_FUNC, data = data)
em = emmeans(lm.fit, ~TREATMENT_SHORT|predicted_endotype)
pairs(em, rev = TRUE) %>% rbind(adjust = "holm")
```

Get confidence intervals.

```{r}
res = summary(pairs(em, rev = TRUE) %>% rbind(adjust = "holm"))

# Structural damage
str.estimate.3 = res$estimate[1]
se = res$SE[1]
df = res$df[1]
str.lower.3 = str.estimate.3 - qt(p = 0.975, df = df)*se
str.upper.3 = str.estimate.3 + qt(p = 0.975, df = df)*se

# Inflammatory
infl.estimate.3 = res$estimate[2]
se = res$SE[2]
df = res$df[2]
infl.lower.3 = infl.estimate.3 - qt(p = 0.975, df = df)*se
infl.upper.3 = infl.estimate.3 + qt(p = 0.975, df = df)*se

# Low turnover
low.estimate.3 = res$estimate[3]
se = res$SE[3]
df = res$df[3]
low.lower.3 = low.estimate.3 - qt(p = 0.975, df = df)*se
low.upper.3 = low.estimate.3 + qt(p = 0.975, df = df)*se
```

Overall test if there is difference in effect (remove interaction term in simpler model).

```{r}
lm.fit2 = lm(WOMACFUNC_CH ~ TREATMENT_SHORT + Age + Sex + BMI + VISIT2_WOMAC_FUNC, data = data)
anova(lm.fit2, lm.fit)
```

Overall effect.

```{r}
em = emmeans(lm.fit2, ~TREATMENT_SHORT)
pairs(em, rev = TRUE) %>% rbind(adjust = "holm")
```

```{r}
res = summary(pairs(em, rev = TRUE) %>% rbind(adjust = "holm"))

# Overall effect
overall.estimate.3 = res$estimate
se = res$SE
df = res$df
overall.lower.3 = overall.estimate.3 - qt(p = 0.975, df = df)*se
overall.upper.3 = overall.estimate.3 + qt(p = 0.975, df = df)*se
```

```{r, fig.width = 5}
dat.forest = tibble(mean = c(str.estimate.3, infl.estimate.3, low.estimate.3, overall.estimate.3),
       lower = c(str.lower.3, infl.lower.3, low.lower.3, overall.lower.3),
       upper = c(str.upper.3, infl.upper.3, low.upper.3, overall.upper.3),
       endotype = c("Structural damage", "Inflammatory", "Low tissue turnover", "Overall"))

dat.forest %>%
  forestplot(labeltext = endotype,
             xlab = "Delta WOMAC Function (%)",
             ci.vertices = TRUE,
             boxsize = 0.25,
             lineheight =  unit(8.6,"mm"),
             col = fpColors(zero = "black")) |>
  fp_add_header(endotype = "Endotype") |>
  fp_set_style(box = "darkblue",
               line = "royalblue",
               txt_gp = fpTxtGp(xlab = gpar(cex = 1.5),
                                ticks = gpar(cex = 1.5),
                                cex = 1.5))
```
 


## WOMAC Stiffness

```{r}
lm.fit = lm(WOMACSTIFF_CH ~ TREATMENT_SHORT + Age + Sex + BMI + TREATMENT_SHORT*predicted_endotype + VISIT2_WOMAC_STIFF, data = data)
em = emmeans(lm.fit, ~TREATMENT_SHORT|predicted_endotype)
pairs(em, rev = TRUE) %>% rbind(adjust = "holm")
```

Get confidence intervals.

```{r}
res = summary(pairs(em, rev = TRUE) %>% rbind(adjust = "holm"))

# Structural damage
str.estimate.4 = res$estimate[1]
se = res$SE[1]
df = res$df[1]
str.lower.4 = str.estimate.4 - qt(p = 0.975, df = df)*se
str.upper.4 = str.estimate.4 + qt(p = 0.975, df = df)*se

# Inflammatory
infl.estimate.4 = res$estimate[2]
se = res$SE[2]
df = res$df[2]
infl.lower.4 = infl.estimate.4 - qt(p = 0.975, df = df)*se
infl.upper.4 = infl.estimate.4 + qt(p = 0.975, df = df)*se

# Low turnover
low.estimate.4 = res$estimate[3]
se = res$SE[3]
df = res$df[3]
low.lower.4 = low.estimate.4 - qt(p = 0.975, df = df)*se
low.upper.4 = low.estimate.4 + qt(p = 0.975, df = df)*se
```

Overall test if there is difference in effect (remove interaction term in simpler model).

```{r}
lm.fit2 = lm(WOMACSTIFF_CH ~ TREATMENT_SHORT + Age + Sex + BMI + VISIT2_WOMAC_STIFF, data = data)
anova(lm.fit2, lm.fit)
```

Overall effect.

```{r}
em = emmeans(lm.fit2, ~TREATMENT_SHORT)
pairs(em, rev = TRUE) %>% rbind(adjust = "holm")
```

```{r}
res = summary(pairs(em, rev = TRUE) %>% rbind(adjust = "holm"))

# Overall effect
overall.estimate.4 = res$estimate
se = res$SE
df = res$df
overall.lower.4 = overall.estimate.4 - qt(p = 0.975, df = df)*se
overall.upper.4 = overall.estimate.4 + qt(p = 0.975, df = df)*se
```

```{r, fig.width = 5}
dat.forest = tibble(mean = c(str.estimate.4, infl.estimate.4, low.estimate.4, overall.estimate.4),
       lower = c(str.lower.4, infl.lower.4, low.lower.4, overall.lower.4),
       upper = c(str.upper.4, infl.upper.4, low.upper.4, overall.upper.4),
       endotype = c("Structural damage", "Inflammatory", "Low tissue turnover", "Overall"))

dat.forest %>%
  forestplot(labeltext = endotype,
             xlab = "Delta WOMAC Stiffness (%)",
             ci.vertices = TRUE,
             boxsize = 0.25, 
             lineheight =  unit(8.6,"mm"),
             col = fpColors(zero = "black")) |>
  fp_add_header(endotype = "Endotype") |>
  fp_set_style(box = "darkblue",
               line = "royalblue",
               txt_gp = fpTxtGp(xlab = gpar(cex = 1.5),
                                ticks = gpar(cex = 1.5),
                                cex = 1.5))
```



## sCTX-I

```{r}
lm.fit = lm(SCTX_CH ~ TREATMENT_SHORT + Age + Sex + BMI + TREATMENT_SHORT*predicted_endotype + VISIT2_sCTX_I, data = data)
em = emmeans(lm.fit, ~TREATMENT_SHORT|predicted_endotype)
pairs(em, rev = TRUE) %>% rbind(adjust = "holm")
```

Get confidence intervals.

```{r}
res = summary(pairs(em, rev = TRUE) %>% rbind(adjust = "holm"))

# Structural damage
str.estimate.5 = res$estimate[1]
se = res$SE[1]
df = res$df[1]
str.lower.5 = str.estimate.5 - qt(p = 0.975, df = df)*se
str.upper.5 = str.estimate.5 + qt(p = 0.975, df = df)*se

# Inflammatory
infl.estimate.5 = res$estimate[2]
se = res$SE[2]
df = res$df[2]
infl.lower.5 = infl.estimate.5 - qt(p = 0.975, df = df)*se
infl.upper.5 = infl.estimate.5 + qt(p = 0.975, df = df)*se

# Low turnover
low.estimate.5 = res$estimate[3]
se = res$SE[3]
df = res$df[3]
low.lower.5 = low.estimate.5 - qt(p = 0.975, df = df)*se
low.upper.5 = low.estimate.5 + qt(p = 0.975, df = df)*se
```

Overall test if there is difference in effect (remove interaction term in simpler model).

```{r}
lm.fit2 = lm(SCTX_CH ~ TREATMENT_SHORT + Age + Sex + BMI + VISIT2_sCTX_I, data = data)
anova(lm.fit2, lm.fit)
```

Overall effect.

```{r}
em = emmeans(lm.fit2, ~TREATMENT_SHORT)
pairs(em, rev = TRUE) %>% rbind(adjust = "holm")
```

```{r}
res = summary(pairs(em, rev = TRUE) %>% rbind(adjust = "holm"))

# Overall effect
overall.estimate.5 = res$estimate
se = res$SE
df = res$df
overall.lower.5 = overall.estimate.5 - qt(p = 0.975, df = df)*se
overall.upper.5 = overall.estimate.5 + qt(p = 0.975, df = df)*se
```

```{r, fig.width = 5}
dat.forest = tibble(mean = c(str.estimate.5, infl.estimate.5, low.estimate.5, overall.estimate.5),
       lower = c(str.lower.5, infl.lower.5, low.lower.5, overall.lower.5),
       upper = c(str.upper.5, infl.upper.5, low.upper.5, overall.upper.5),
       endotype = c("Structural damage", "Inflammatory", "Low tissue turnover", "Overall"))

dat.forest %>%
  forestplot(labeltext = endotype,
             xlab = "Delta sCTX-I (%)",
             ci.vertices = TRUE,
             boxsize = 0.25,
             lineheight =  unit(8.6,"mm"),
             col = fpColors(zero = "black")) |>
  fp_add_header(endotype = "Endotype") |>
  fp_set_style(box = "darkblue",
               line = "royalblue",
               txt_gp = fpTxtGp(xlab = gpar(cex = 1.5),
                                ticks = gpar(cex = 1.5),
                                cex = 1.5))
```



## uCTX-II

```{r}
lm.fit = lm(UCTXII_CH ~ TREATMENT_SHORT + Age + Sex + BMI + TREATMENT_SHORT*predicted_endotype + VISIT2_uCTX_II_CREA, data = data)
em = emmeans(lm.fit, ~TREATMENT_SHORT|predicted_endotype)
pairs(em, rev = TRUE) %>% rbind(adjust = "holm")
```

Get confidence intervals.

```{r}
res = summary(pairs(em, rev = TRUE) %>% rbind(adjust = "holm"))

# Structural damage
str.estimate.6 = res$estimate[1]
se = res$SE[1]
df = res$df[1]
str.lower.6 = str.estimate.6 - qt(p = 0.975, df = df)*se
str.upper.6 = str.estimate.6 + qt(p = 0.975, df = df)*se

# Inflammatory
infl.estimate.6 = res$estimate[2]
se = res$SE[2]
df = res$df[2]
infl.lower.6 = infl.estimate.6 - qt(p = 0.975, df = df)*se
infl.upper.6 = infl.estimate.6 + qt(p = 0.975, df = df)*se

# Low turnover
low.estimate.6 = res$estimate[3]
se = res$SE[3]
df = res$df[3]
low.lower.6 = low.estimate.6 - qt(p = 0.975, df = df)*se
low.upper.6 = low.estimate.6 + qt(p = 0.975, df = df)*se
```

Overall test if there is difference in effect (remove interaction term in simpler model).

```{r}
lm.fit2 = lm(UCTXII_CH ~ TREATMENT_SHORT + Age + Sex + BMI + VISIT2_uCTX_II_CREA, data = data)
anova(lm.fit2, lm.fit)
```

Overall effect.

```{r}
em = emmeans(lm.fit2, ~TREATMENT_SHORT)
pairs(em, rev = TRUE) %>% rbind(adjust = "holm")
```

```{r}
res = summary(pairs(em, rev = TRUE) %>% rbind(adjust = "holm"))

# Overall effect
overall.estimate.6 = res$estimate
se = res$SE
df = res$df
overall.lower.6 = overall.estimate.6 - qt(p = 0.975, df = df)*se
overall.upper.6 = overall.estimate.6 + qt(p = 0.975, df = df)*se
```

```{r, fig.width = 5}
dat.forest = tibble(mean = c(str.estimate.6, infl.estimate.6, low.estimate.6, overall.estimate.6),
       lower = c(str.lower.6, infl.lower.6, low.lower.6, overall.lower.6),
       upper = c(str.upper.6, infl.upper.6, low.upper.6, overall.upper.6),
       endotype = c("Structural damage", "Inflammatory", "Low tissue turnover", "Overall"))

dat.forest %>%
  forestplot(labeltext = endotype,
             xlab = "Delta uCTX-II (%)",
             ci.vertices = TRUE,
             boxsize = 0.25,
             lineheight =  unit(8.6,"mm"),
             col = fpColors(zero = "black")) |>
  fp_add_header(endotype = "Endotype") |>
  fp_set_style(box = "darkblue",
               line = "royalblue",
               txt_gp = fpTxtGp(xlab = gpar(cex = 1.5),
                                ticks = gpar(cex = 1.5),
                                cex = 1.5))
```



## sCTX-I - numerical difference

```{r}
lm.fit = lm(SCTX_CH_NUM ~ TREATMENT_SHORT + Age + Sex + BMI + TREATMENT_SHORT*predicted_endotype + VISIT2_sCTX_I, data = data)
em = emmeans(lm.fit, ~TREATMENT_SHORT|predicted_endotype)
pairs(em, rev = TRUE) %>% rbind(adjust = "holm")
```

Get confidence intervals.

```{r}
res = summary(pairs(em, rev = TRUE) %>% rbind(adjust = "holm"))

# Structural damage
str.estimate.7 = res$estimate[1]
se = res$SE[1]
df = res$df[1]
str.lower.7 = str.estimate.7 - qt(p = 0.975, df = df)*se
str.upper.7 = str.estimate.7 + qt(p = 0.975, df = df)*se

# Inflammatory
infl.estimate.7 = res$estimate[2]
se = res$SE[2]
df = res$df[2]
infl.lower.7 = infl.estimate.7 - qt(p = 0.975, df = df)*se
infl.upper.7 = infl.estimate.7 + qt(p = 0.975, df = df)*se

# Low turnover
low.estimate.7 = res$estimate[3]
se = res$SE[3]
df = res$df[3]
low.lower.7 = low.estimate.7 - qt(p = 0.975, df = df)*se
low.upper.7 = low.estimate.7 + qt(p = 0.975, df = df)*se
```

Overall test if there is difference in effect (remove interaction term in simpler model).

```{r}
lm.fit2 = lm(SCTX_CH_NUM ~ TREATMENT_SHORT + Age + Sex + BMI + VISIT2_sCTX_I, data = data)
anova(lm.fit2, lm.fit)
```

Overall effect.

```{r}
em = emmeans(lm.fit2, ~TREATMENT_SHORT)
pairs(em, rev = TRUE) %>% rbind(adjust = "holm")
```

```{r}
res = summary(pairs(em, rev = TRUE) %>% rbind(adjust = "holm"))

# Overall effect
overall.estimate.7 = res$estimate
se = res$SE
df = res$df
overall.lower.7 = overall.estimate.7 - qt(p = 0.975, df = df)*se
overall.upper.7 = overall.estimate.7 + qt(p = 0.975, df = df)*se
```

```{r, fig.width = 5}
dat.forest = tibble(mean = c(str.estimate.7, infl.estimate.7, low.estimate.7, overall.estimate.7),
       lower = c(str.lower.7, infl.lower.7, low.lower.7, overall.lower.7),
       upper = c(str.upper.7, infl.upper.7, low.upper.7, overall.upper.7),
       endotype = c("Structural damage", "Inflammatory", "Low tissue turnover", "Overall"))

dat.forest %>%
  forestplot(labeltext = endotype,
             xlab = "Delta sCTX-I (num)",
             ci.vertices = TRUE,
             boxsize = 0.25,
             lineheight =  unit(8.6,"mm"),
             col = fpColors(zero = "black")) |>
  fp_add_header(endotype = "Endotype") |>
  fp_set_style(box = "darkblue",
               line = "royalblue",
               txt_gp = fpTxtGp(xlab = gpar(cex = 1.5),
                                ticks = gpar(cex = 1.5),
                                cex = 1.5))
```



## uCTX-II - numerical difference

```{r}
lm.fit = lm(UCTXII_CH_NUM ~ TREATMENT_SHORT + Age + Sex + BMI + TREATMENT_SHORT*predicted_endotype + VISIT2_uCTX_II_CREA, data = data)
em = emmeans(lm.fit, ~TREATMENT_SHORT|predicted_endotype)
pairs(em, rev = TRUE) %>% rbind(adjust = "holm")
```

Get confidence intervals.

```{r}
res = summary(pairs(em, rev = TRUE) %>% rbind(adjust = "holm"))

# Structural damage
str.estimate.8 = res$estimate[1]
se = res$SE[1]
df = res$df[1]
str.lower.8 = str.estimate.8 - qt(p = 0.975, df = df)*se
str.upper.8 = str.estimate.8 + qt(p = 0.975, df = df)*se

# Inflammatory
infl.estimate.8 = res$estimate[2]
se = res$SE[2]
df = res$df[2]
infl.lower.8 = infl.estimate.8 - qt(p = 0.975, df = df)*se
infl.upper.8 = infl.estimate.8 + qt(p = 0.975, df = df)*se

# Low turnover
low.estimate.8 = res$estimate[3]
se = res$SE[3]
df = res$df[3]
low.lower.8 = low.estimate.8 - qt(p = 0.975, df = df)*se
low.upper.8 = low.estimate.8 + qt(p = 0.975, df = df)*se
```

Overall test if there is difference in effect (remove interaction term in simpler model).

```{r}
lm.fit2 = lm(UCTXII_CH_NUM ~ TREATMENT_SHORT + Age + Sex + BMI + VISIT2_uCTX_II_CREA, data = data)
anova(lm.fit2, lm.fit)
```

Overall effect.

```{r}
em = emmeans(lm.fit2, ~TREATMENT_SHORT)
pairs(em, rev = TRUE) %>% rbind(adjust = "holm")
```

```{r}
res = summary(pairs(em, rev = TRUE) %>% rbind(adjust = "holm"))

# Overall effect
overall.estimate.8 = res$estimate
se = res$SE
df = res$df
overall.lower.8 = overall.estimate.8 - qt(p = 0.975, df = df)*se
overall.upper.8 = overall.estimate.8 + qt(p = 0.975, df = df)*se
```

```{r, fig.width = 5}
dat.forest = tibble(mean = c(str.estimate.8, infl.estimate.8, low.estimate.8, overall.estimate.8),
       lower = c(str.lower.8, infl.lower.8, low.lower.8, overall.lower.8),
       upper = c(str.upper.8, infl.upper.8, low.upper.8, overall.upper.8),
       endotype = c("Structural damage", "Inflammatory", "Low tissue turnover", "Overall"))

dat.forest %>%
  forestplot(labeltext = endotype,
             xlab = "Delta uCTX-II (num)",
             ci.vertices = TRUE,
             boxsize = 0.25,
             lineheight =  unit(8.6,"mm"),
             col = fpColors(zero = "black")) |>
  fp_add_header(endotype = "Endotype") |>
  fp_set_style(box = "darkblue",
               line = "royalblue",
               txt_gp = fpTxtGp(xlab = gpar(cex = 1.5),
                                ticks = gpar(cex = 1.5),
                                cex = 1.5))
```


# Collect results

```{r, echo = FALSE}
tibble(endotype = c("structural damage", "inflammatory", "low turnover", "overall"),
       delta_NRSKNEEPAIN = c(paste0(format(round(str.estimate.1, 3), nsmall = 3), " (", format(round(str.lower.1, 3), nsmall = 3), ",", format(round(str.upper.1, 3), nsmall = 3), ")"),
                paste0(format(round(infl.estimate.1, 3), nsmall = 3), " (", format(round(infl.lower.1, 3), nsmall = 3), ",", format(round(infl.upper.1, 3), nsmall = 3), ")"),
                paste0(format(round(low.estimate.1, 3), nsmall = 3), " (", format(round(low.lower.1, 3), nsmall = 3), ",", format(round(low.upper.1, 3), nsmall = 3), ")"),
                paste0(format(round(overall.estimate.1, 3), nsmall = 3), " (", format(round(overall.lower.1, 3), nsmall = 3), ",", format(round(overall.upper.1, 3), nsmall = 3), ")")),
       delta_WOMACPAIN_pct = c(paste0(format(round(str.estimate.2, 2), nsmall = 2), " (", format(round(str.lower.2, 2), nsmall = 2), ",", format(round(str.upper.2, 2), nsmall = 2), ")"),
                paste0(format(round(infl.estimate.2, 2), nsmall = 2), " (", format(round(infl.lower.2, 2), nsmall = 2), ",", format(round(infl.upper.2, 2), nsmall = 2), ")"),
                paste0(format(round(low.estimate.2, 2), nsmall = 2), " (", format(round(low.lower.2, 2), nsmall = 2), ",", format(round(low.upper.2, 2), nsmall = 2), ")"),
                paste0(format(round(overall.estimate.2, 2), nsmall = 2), " (", format(round(overall.lower.2, 2), nsmall = 2), ",", format(round(overall.upper.2, 2), nsmall = 2), ")")),
       delta_WOMACFUNC_pct = c(paste0(format(round(str.estimate.3, 2), nsmall = 2), " (", format(round(str.lower.3, 2), nsmall = 2), ",", format(round(str.upper.3, 3), nsmall = 3), ")"),
                paste0(format(round(infl.estimate.3, 2), nsmall = 2), " (", format(round(infl.lower.3, 2), nsmall = 2), ",", format(round(infl.upper.3, 2), nsmall = 2), ")"),
                paste0(format(round(low.estimate.3, 2), nsmall = 2), " (", format(round(low.lower.3, 2), nsmall = 2), ",", format(round(low.upper.3, 2), nsmall = 2), ")"),
                paste0(format(round(overall.estimate.3, 2), nsmall = 2), " (", format(round(overall.lower.3, 2), nsmall = 2), ",", format(round(overall.upper.3, 2), nsmall = 2), ")")),
       delta_WOMACSTIFF_pct = c(paste0(format(round(str.estimate.4, 2), nsmall = 2), " (", format(round(str.lower.4, 2), nsmall = 2), ",", format(round(str.upper.4, 3), nsmall = 3), ")"),
                paste0(format(round(infl.estimate.4, 2), nsmall = 2), " (", format(round(infl.lower.4, 2), nsmall = 2), ",", format(round(infl.upper.4, 2), nsmall = 2), ")"),
                paste0(format(round(low.estimate.4, 2), nsmall = 2), " (", format(round(low.lower.4, 2), nsmall = 2), ",", format(round(low.upper.4, 2), nsmall = 2), ")"),
                paste0(format(round(overall.estimate.4, 2), nsmall = 2), " (", format(round(overall.lower.4, 2), nsmall = 2), ",", format(round(overall.upper.4, 2), nsmall = 2), ")")),
       delta_sCTXI_pct = c(paste0(format(round(str.estimate.5, 2), nsmall = 2), " (", format(round(str.lower.5, 2), nsmall = 2), ",", format(round(str.upper.5, 2), nsmall = 2), ")"),
                paste0(format(round(infl.estimate.5, 2), nsmall = 2), " (", format(round(infl.lower.5, 2), nsmall = 2), ",", format(round(infl.upper.5, 2), nsmall = 2), ")"),
                paste0(format(round(low.estimate.5, 2), nsmall = 2), " (", format(round(low.lower.5, 2), nsmall = 2), ",", format(round(low.upper.5, 2), nsmall = 2), ")"),
                paste0(format(round(overall.estimate.5, 2), nsmall = 2), " (", format(round(overall.lower.5, 2), nsmall = 2), ",", format(round(overall.upper.5, 2), nsmall = 2), ")")),
       delta_uCTX_II_pct = c(paste0(format(round(str.estimate.6, 2), nsmall = 2), " (", format(round(str.lower.6, 2), nsmall = 2), ",", format(round(str.upper.6, 2), nsmall = 2), ")"),
                paste0(format(round(infl.estimate.6, 2), nsmall = 2), " (", format(round(infl.lower.6, 2), nsmall = 2), ",", format(round(infl.upper.6, 2), nsmall = 2), ")"),
                paste0(format(round(low.estimate.6, 2), nsmall = 2), " (", format(round(low.lower.6, 2), nsmall = 2), ",", format(round(low.upper.6, 2), nsmall = 2), ")"),
                paste0(format(round(overall.estimate.6, 2), nsmall = 2), " (", format(round(overall.lower.6, 2), nsmall = 2), ",", format(round(overall.upper.6, 2), nsmall = 2), ")")),
       delta_sCTXI_num = c(paste0(format(round(str.estimate.7, 2), nsmall = 2), " (", format(round(str.lower.7, 2), nsmall = 2), ",", format(round(str.upper.7, 2), nsmall = 2), ")"),
                paste0(format(round(infl.estimate.7, 2), nsmall = 2), " (", format(round(infl.lower.7, 2), nsmall = 2), ",", format(round(infl.upper.7, 2), nsmall = 2), ")"),
                paste0(format(round(low.estimate.7, 2), nsmall = 2), " (", format(round(low.lower.7, 2), nsmall = 2), ",", format(round(low.upper.7, 2), nsmall = 2), ")"),
                paste0(format(round(overall.estimate.7, 2), nsmall = 2), " (", format(round(overall.lower.7, 2), nsmall = 2), ",", format(round(overall.upper.7, 2), nsmall = 2), ")")),
       delta_uCTX_II_num = c(paste0(format(round(str.estimate.8, 2), nsmall = 2), " (", format(round(str.lower.8, 2), nsmall = 2), ",", format(round(str.upper.8, 2), nsmall = 2), ")"),
                paste0(format(round(infl.estimate.8, 2), nsmall = 2), " (", format(round(infl.lower.8, 2), nsmall = 2), ",", format(round(infl.upper.8, 2), nsmall = 2), ")"),
                paste0(format(round(low.estimate.8, 2), nsmall = 2), " (", format(round(low.lower.8, 2), nsmall = 2), ",", format(round(low.upper.8, 2), nsmall = 2), ")"),
                paste0(format(round(overall.estimate.8, 2), nsmall = 2), " (", format(round(overall.lower.8, 2), nsmall = 2), ",", format(round(overall.upper.8, 2), nsmall = 2), ")"))) %>%
  pivot_longer(cols = !contains("endotype"),
               values_to = "effect_size_95_CI",
               names_to = "parameter") %>%
  arrange(parameter, endotype) %>%
  as.data.frame
```

Mean change in parameters split by treatment and endotype.

```{r}
data %>%
  group_by(predicted_endotype, TREATMENT_SHORT) %>%
  summarize(mean_NRSKNEEPAIN_CH = mean(NRSKNEEPAIN_CH, na.rm = TRUE) %>% round(digits = 3) %>% format(nsmall = 3),
            mean_WOMACPAIN_CH = mean(WOMACPAIN_CH, na.rm = TRUE) %>% round(digits = 2) %>% format(nsmall = 2),
            mean_WOMACFUNC_CH = mean(WOMACFUNC_CH, na.rm = TRUE) %>% round(digits = 2) %>% format(nsmall = 2),
            mean_WOMACSTIFF_CH = mean(WOMACSTIFF_CH, na.rm = TRUE) %>% round(digits = 2) %>% format(nsmall = 2),
            mean_SCTXI_CH = mean(SCTX_CH, na.rm = TRUE) %>% round(digits = 2) %>% format(nsmall = 2),
            mean_UCTXII_CH = mean(UCTXII_CH, na.rm = TRUE) %>% round(digits = 2) %>% format(nsmall = 2),
            mean_SCTXI_CH_NUM = mean(SCTX_CH_NUM, na.rm = TRUE) %>% round(digits = 2) %>% format(nsmall = 2),
            mean_UCTXII_CH_NUM = mean(UCTXII_CH_NUM, na.rm = TRUE) %>% round(digits = 2) %>% format(nsmall = 2)) %>%
  pivot_longer(contains("mean"),
               names_to = "parameter",
               values_to = "mean_change") %>%
  arrange(parameter, predicted_endotype) %>%
  as.data.frame
```

Mean change in parameters split by treatment on overall population.

```{r}
data %>%
  group_by(TREATMENT_SHORT) %>%
  summarize(mean_NRSKNEEPAIN_CH = mean(NRSKNEEPAIN_CH, na.rm = TRUE) %>% round(digits = 3) %>% format(nsmall = 3),
            mean_WOMACPAIN_CH = mean(WOMACPAIN_CH, na.rm = TRUE) %>% round(digits = 2) %>% format(nsmall = 2),
            mean_WOMACFUNC_CH = mean(WOMACFUNC_CH, na.rm = TRUE) %>% round(digits = 2) %>% format(nsmall = 2),
            mean_WOMACSTIFF_CH = mean(WOMACSTIFF_CH, na.rm = TRUE) %>% round(digits = 2) %>% format(nsmall = 2),
            mean_SCTXI_CH = mean(SCTX_CH, na.rm = TRUE) %>% round(digits = 2) %>% format(nsmall = 2),
            mean_UCTXII_CH = mean(UCTXII_CH, na.rm = TRUE) %>% round(digits = 2) %>% format(nsmall = 2),
            mean_SCTXI_CH_NUM = mean(SCTX_CH_NUM, na.rm = TRUE) %>% round(digits = 2) %>% format(nsmall = 2),
            mean_UCTXII_CH_NUM = mean(UCTXII_CH_NUM, na.rm = TRUE) %>% round(digits = 2) %>% format(nsmall = 2)) %>%
  pivot_longer(contains("mean"),
               names_to = "parameter",
               values_to = "mean_change") %>%
  arrange(parameter) %>%
  as.data.frame
```
